Graph Topological Aspects of Granger Causal Network Learning

### 2019-07-18 paper_theory/paper_theory.tex
Better proofs.

@@

\abstract{We study Granger-causality in the context of wide-sense
  stationary time series, where our focus is on the topological
  aspects of the underlying causality graph.  We establish sufficient
  conditions (in particular, we develop the notion of a ``strongly
  causal'' graph topology) under which the true causality graph can be
  recovered via pairwise causality testing alone, and provide examples
  from the gene regulatory network literature suggesting that our
  concept of a strongly causal graph may be applicable to this field.}

@@

% \clearpage

% % \tableofcontents
% \clearpage

\section{Introduction and Review}
\label{sec:introduction}
In this paper we study the notion of Granger-causality
\cite{granger1969investigating} \cite{Granger1980329} as a means of
uncovering an underlying causal structure in multivariate time series.
Though the underlying causality graph cannot be observed directly, we
will infer it's presence as a latent structure among our observed time
series data.  This notion is leveraged in a variety of applications
e.g. in Neuroscience as a means of recovering interactions amongst
brain regions \cite{bressler2011wiener}, \cite{anna_paper2008},
\cite{david2008identifying}; in the study of the dependence and
connectedness of financial institutions \cite{NBERw16223}; gene
expression networks \cite{Fujita2007},
\cite{methods_for_inferring_gene_regulatory_networks_from_time_series_expression_data},
\cite{grouped_graphical_granger_modelling_for_gene_expression_regulatory_networks_discovery},
\cite{discovering_graphical_Granger_causality_using_the_truncating_lasso_penalty};
and power system design \cite{Misyrlis2016450}, \cite{yuan2014root}.

Granger-causality can generally be formulated by searching for the
``best'' graph structure consistent with observed data, which is in
general an extremely challenging problem (i.e. it may be framed as a
best subset selection problem, see \cite{bss_mio},
\cite{hastie_bss_comp}), moreover, the comparison of quality between
different structures, and hence the notion of ``best'' needs
qualification.  In applications where we are interested merely in
minimizing the mean squared error of a linear one-step-ahead
predictor, then we will be satisfied with an entirely dense graph of
connections, since each edge can only serve to reduce estimation
error.  However, since the number of edges scales quadratically in $n$
(the number of nodes) it becomes imperative to infer a sparse
causality graph for large systems, both to avoid overfitting observed
data, as well as to aid the interpretability of the results.

A fairly early approach to the problem in the context of large systems
is provided by \cite{bach2004learning}, where the authors apply a
local search heuristic to the Whittle likelihood with an AIC
penalization.  The local search heuristic where at each iteration an
edge is either added, removed, or reversed is a common approach to
combinatorial optimization due to it's simplicity, but is liable to
get stuck in shallow local minima.

A second and wildly successful heuristic is the LASSO regularizer
\cite{tibshirani1996regression}, which can be understood as a natural
convex relaxation to penalizing the count of the non-zero edges.  The
LASSO enjoys fairly strong theoretical guarantees
\cite{wainwright2009sharp}, extending largely to the case of
stationary time series data with a sufficiently fast rate of
dependence decay \cite{basu2015} \cite{wong2016lasso}
\cite{autoregressive_process_modelling_via_the_lasso_procedure}, and
variations on the LASSO have been applied in a number of different
time series contexts as well as Granger-causality
\cite{DBLP:journals/corr/HallacPBL17} \cite{haufe2008sparse}
\cite{bolstad2011causal} \cite{he2013stationary}
\cite{grouped_graphical_granger_modelling_for_gene_expression_regulatory_networks_discovery}.
One of the key improvements to the original LASSO algorithm is the
adaptive (i.e. weighted) ``adaLASSO'' \cite{adaptive_lasso_zou2006},
for which oracle results (i.e. asymptotic support recovery) are
established under less restrictive conditions than for the vanilla
LASSO.  We study a novel perspective on the problem by establishing
conditions under which simple \textit{pairwise} Granger-causality
tests are sufficient to recover the entire graph.

%   Prior assumptions
% on the \textit{topology} of the underlying Granger-causality graph can
% result in tests more powerful than more generally applicable
% algorithms, and we provide examples from the literature where
% simplifying topological assumptions may be met.

% \subsection{Contributions}
In the context of time series data, sparsity assumptions remain
important, but there is significant additional structure that may
arise as a result of considering the topology of the underlying
Granger-causality graph, which to our knowledge remains largely
unexplored.  The focus of this paper is to shed light on some of these
topological questions, in particular, we study a particularly simple
notion of causality graph topology which we term ``strongly causal''
and show that stationary times series whose underlying causality graph
has this structure satisfy natural intuitive notions of ``information
flow'' through the graph.  Moreover, we show that such graphs are
perfectly recoverable with only \textit{pairwise} Granger-causality
tests, which would otherwise suffer from serious confounding problems.
Aside from being an interesting theoretical perspective, prior
assumptions about the underlying graph (similarly to sparsity
assumptions) can greatly improve upon the statistical power of
causality graph recovery algorithms when the assumptions are met.

@@

The principle contributions of this paper are as follows: firstly, in
section \ref{sec:theory} we study \textit{pairwise} Granger-causality
relations, providing novel theorems connecting the structure of the
causality graph to the pairwise ``causality flow'' in the system, as
well as an interpretation in terms of the graph topology of the
sparsity pattern of matrices arising in the Wold decomposition,
generalizing in some sense the notion of ``feedback-free'' processes
studied by \cite{caines1975feedback} in close connection with
Granger-causality.  We establish sufficient conditions (sections
\ref{sec:strongly_causal_graphs}, \ref{sec:persistent_systems}) under
which a fully conditional Granger-causality graph can be recovered
from pairwise tests alone (sec \ref{sec:pairwise_algorithm}).  We
report a summary of simulation results in \ref{sec:simulation}, with
additional results reported in the supplementary material
\ref{apx:simulation}.  Our simulation results establish that there is
significant potential for improvement over existing methods, and that
the graph-topological aspects of time series analysis are relevant for
both theory and practice.  Concluding remarks on further open problems
and extentions are provided in Section \ref{sec:conclusion}.  The
proofs of each proposition and theorem are also relegated to the
supplementary material, simple corollaries have proofs included in the
main text.

@@

\begin{equation}
  \label{eq:hilbert_space_defn}
  \begin{aligned}
    % \H^x &= \cl \{\sum_{\tau = -T}^T a_\tau^\T x(t - \tau)\ |\ a_\tau \in \R^n, T \in \N\} \subseteq L_2(\Omega),\\
    \H_t^x &= \cl \{\sum_{\tau = 0}^p a_\tau^\T x(t - \tau)\ |\ a_\tau \in \R^n, p \in \N\} \subseteq L_2(\Omega)\\
    H_t^x &= \{a x(t)\ |\ a \in \R\} \subseteq L_2(\Omega),
  \end{aligned}
\end{equation}

where the closure is naturally in mean-square.  We will often omit the
superscript $x$ which should be clear from context.  Evidently these
spaces are separable, and as closed subspaces of a Hilbert space they
are themselves Hilbert.  We will denote the spaces generated in
analogous ways by particular components of $x$ as e.g.
$\H_t^{(i, j)}$, $\H_t^{i}$ or by all but particular components as
$\H_t^{-j}$.

@@

where $c(t)$ is a purely deterministic sequence$v(t)$ is an
uncorrelated sequence and $A(0) = I$.  We will assume that $c(t) = 0$.
We additionally require that this representation can be inverted to
yield the $\VAR(\infty)$ form

@@

\begin{definition}[Granger Causality]
  \label{def:granger_causality}
  For the WSS series $x(t)$ satisfying the assumptions of Definition
  \ref{def:basic_setup} we will say that component $x_j$
  \textit{Granger-Causes} (GC) component $x_i$ (with respect to $x$)
  and write $x_j \gc x_i$ if given Hilbert spaces $\H_{t - 1}$,
  $\H^{-j}_{t - 1}$

@@

where $\xi[x \ |\ \H] = \E (x - \linE{x}{\H})^2$ is the mean squared
estimation error and $\linE{x}{\H} = \text{proj}_{\H}(x)$ denotes the
(unique) projection onto the Hilbert space $\H$.
\end{definition}

@@

% \begin{proof}
%   % Geweke uses the log of the ratio of the determinants of the residual variances
%   % 
%   % The equivalence $(1) \iff (3)$ is essentially a restatement of the
%   % definition, but is in line with the seminal work of Geweke
%   % \cite{geweke1982measurement}, \cite{geweke1984}.

%   $(a) \Rightarrow (b)$ follows as a result of the uniqueness of orthogonal
%   projection (i.e. the best estimate is necessarily the coefficients
%   of the model).  $(b) \Rightarrow (c)$ follows since in computing
%   $(y - \linE{y}{\H_{t - 1}^{-j}})$ for $y \in H_t^i$ it is sufficient
%   to consider $y = x_i(t)$ by linearity, then since
%   $H_{t - 1}^i \subseteq \H_{t - 1}^{-j}$ we have
%   $(x_i(t) - \linE{x_i(t)}{\H_{t = 1}^{-j}}) = v_i(t)$ since
%   $\B_{ij}(z) = 0$ and $v_i(t) \perp \H_{t - 1}$.  $(c) \iff (d)$ is a result
%   of the equivalence in Definition
%   \ref{lem:conditional_orthogonality_equivalence}.  And,
%   $(d) \implies (a)$ follows directly from the Definition.
% \end{proof}

@@

  A \textit{cycle} is a path of length $2$ or more between a node and
  itself.  An edge between a node and itself $(i, i) \in \gcge$ (which
  is not a cycle) is referred to as a \textit{loop}.

@@

  Recall that we do not allow a node to be it's own parent, however
  unless $\gcg$ is a DAG a node \textit{can} be it's own ancestor.  We will
  ocassionally need to explicitly exclude $i$ from $\anc{i}$, in which
  case we will write $\anc{i}\setminus \{i\}$.
\end{definition}

@@

The following propositions are essentially lemmas used for the proof
of the upcomming Proposition \ref{prop:ancestor_properties}, but
remain relevant for providing intuitive insight into the problems at
hand.

@@

  then $\H_t^{(i)} \perp \H_t^{(j)}$, that is,
  $\forall s, \tau \in \Z_+\ \E[x_i(t - s)x_j(t - \tau)] = 0$.  Moreover,
  this means that $j \npwgc i$ and $\linE{x_j(t)}{\H_{t - 1}^i} = 0$.
\end{proposition}

@@

\begin{remark}
  It is possible for components of $x(t)$ to be correlated at some
  time lags without resulting in pairwise causality.  For instance,
  the conclusion $j \npwgc i$ of Proposition
  \ref{prop:separated_ancestor_uncorrelated} will still hold even if
  $i \in \anc{j}$, since $j$ cannot provide any information about $i$
  that is not available from observing $i$ itself.  However, it is
  important to note that this is the case only if we have direct
  access to $x(t)$.  That is, if we instead observed a noisy version
  of $x$ (i.e. $\tilde{x}(t) = x(t) + n(t)$) the observation of
  $\tilde{x}_j$ would enable conditionally improved estimates of
  $\tilde{x}_i$, without there necessarily being any causal flow in
  the underlying system $x(t)$.
\end{remark}

@@

\begin{definition}[Confounder]
  A node $k$ will be referred to as a \textit{confounder} of nodes
  $i, j$ (neither of which are equal to $k$) if
  $k \in \anc{i} \cap \anc{j}$ and there exists a path
  $\gcgpath{k}{i}$ not containing $j$, and a path $\gcgpath{k}{j}$
  not containing $i$.

  A simple example is furnished by the ``fork'' graph
  $i \leftarrow k \rightarrow j$.
\end{definition}

@@

It seems reasonable to expect a converse of proposition
\ref{prop:ancestor_properties} to hold, i.e.
$j \in \anc{i} \Rightarrow j \pwgc i$.  Unfortunately, this is not the
case in general, as different paths through $\gcg$ can lead to
cancellation (see Figure \ref{fig:diamond_cancellation}).  In fact, we
do not even have $j \in \pa{i} \Rightarrow j \pwgc i$ (see Figure
\ref{fig:lag_cancellation}).

@@

Examples of strongly causal graphs include directed trees (or
forests), DAGs where each node has at most one parent, and figure
\ref{fig:example_fig3} of this paper.  A complete bipartite graph with
$2n$ nodes is also strongly causal, demonstrating that the number of
edges of such a graph can still scale quadratically with the number of
nodes.  It is evident that the strong causal property is inherited by
subgraphs.

\begin{example}
  Though examples of SCGs are easy to construct in theory, should
  practitioners expect SCGs to arise in application?  While a positive
  answer to this question is not \textit{necessary} for the concept to
  be useful, it is certainly sufficient.  Though the answer is likely
  to depend upon the particular application area, examples appear to
  be available in biology, in particular, the authors of
  \cite{discovering_graphical_Granger_causality_using_the_truncating_lasso_penalty}
  cite an example of the so called ``transcription regulatory network
  of \textit{E.coli}'', and
  \cite{learning_genome_scale_regulatory_networks} study a much larger
  regulatory network of \textit{Saccharomyces cerevisiae}.  These
  networks, which we reproduce in figure \ref{fig:gene_networks}
  appear to have at most a small number of edges which violate the
  strong-causality condition.

@@

  {\footnotesize Figure \ref{fig:gene_network1}
    is reproduced under the Creative Commons Attribution
    Non-Commercial License
    (http://creativecommons.org/licenses/by-nc/2.5) and Figure
    \ref{fig:gene_network2} under the Creative Commons Attribution
    License (https://creativecommons.org/licenses/by/4.0/)}
  \end{figure}
\end{example}

For later use, and to get a feel for the topological implications of
strong causality, we explore a number of properties of such graphs
before moving into the main result of this section.  The following
important property essentially strengthens proposition
\ref{prop:ancestor_properties} for the case of strongly causal graphs.

@@

\begin{corollary}
  \label{cor:parent_corollary}
  If $\gcg$ is a strongly causal DAG then $i \pwgc j$ and $j \in \anc{i}$ are
  \textit{alternatives}, that is $i \pwgc j \Rightarrow j \notin \anc{i}$.
\end{corollary}
\begin{proof}
  Suppose that $i \pwgc j$ and $j \in \anc{i}$.  Then since $\gcg$ is
  acyclic $i \not\in \anc{j}$, and by proposition
  \ref{prop:ancestor_properties} there is some
  $k \in \anc{i}\cap\anc{j}$ which is a confounder.  However, by
  proposition \ref{prop:sc_graph_common_anc} $k$ cannot be a
  confounder, a contradiction.
\end{proof}
% This is a direct proof
  % \begin{proof}
%   Suppose that $\gcg$ is a strongly causal DAG and that we have both
%   $i \pwgc j$ and $j \in \anc{i}$, which implies that
%   $i \not \in \anc{j}$ since $\gcg$ is a DAG.  We will establish the
%   contradiction $i \in \anc{j}$.

%   Since $j \in \anc{i}$ there is a path $\gcgpath{j}{i}$.
%   Moreover, since $i \pwgc j$ by proposition \ref{prop:pwgc_anc} there
%   must be a confounding ancestor $k \in \anc{i} \cap \anc{j}$, where the
%   node $k$ has a path to $j,\ \gcgpath{k}{j}$, as well as a path
%   to $i,\ \gcgpath{k}{i}$.

%   \hl{Double check that the ancestor properties proposition implies
%     that BOTH paths do not contain the other node.}

@@

\begin{corollary}
  \label{cor:bidirectional_edge}
  If $\gcg$ is a strongly causal DAG such that $i \pwgc j$ and
  $j \pwgc i$, then $i \not\in \anc{j}$ and $j \not\in \anc{i}$.  In
  particular, a pairwise bidirectional edge indicates the absence of
  any edge in $\gcg$.
\end{corollary}
\begin{proof}
  This follows directly from applying proposition
  \ref{cor:parent_corollary} to $i \pwgc j$ and $j \pwgc i$.
% This is a direct proof
% \begin{proof}
  % By way of contradiction, suppose that $i \in \pa{j}$.  We will
  % consider the two possibilities allowed by proposition
  % \ref{prop:ancestor_properties} for $j \pwgc i$.  Firstly
  % $j \in \anc{i}$ is impossible since $\gcg$ is assumed to be acyclic.
  % Secondly, if there is some confounding node
  % $u \in \anc{i} \cap \anc{j}$ with a path
  % $u \rightarrow \cdots \rightarrow j$ which does not contain $i$ we
  % have a contradiction since there must now be multiple
  % $u \rightarrow j$ paths: the aforementioned, and a path
  % $u \rightarrow \cdots \rightarrow i \rightarrow \cdots \rightarrow
  % j$ which \textit{does} contain $i$.  We conclude that $i \in \pa{j}$
  % is impossible, and symmetrically that $j \in \pa{i}$ is as well.
% \end{proof}
\end{proof}

In light of proposition \ref{prop:sc_graph_common_anc}, the following
provides a partial converse to proposition
\ref{prop:ancestor_properties}, and supports the intuition of ``causal
flow'' through paths in $\gcg$.

@@

\begin{remark}
  As we have seen and as is true in much of statistics, confounding
  nodes pose challenges for Granger-causality.  However, as opposed to
  Pearl's causal calculus \cite{pearl2000art}, pairwise
  Granger-causality does not suffer any difficulty with so-called
  ``colliders'' (in the language of Pearl's causal calculus
  \cite{pearl2009causality}), that is, the topology
  $i \rightarrow k \leftarrow j$ will never result in $i \pwgc j$ or
  $j \pwgc i$.  This is evidently an advantage of the
  \textit{temporal} nature of Granger-causality -- there is no
  backwards causal flow along the edges of $\gcg$.
\end{remark}

\begin{example}
  As a final remark of this subsection we note that a complete
  converse to proposition \ref{prop:ancestor_properties} is not
  possible without additional conditions.  Consider the ``fork''
  system on $3$ nodes (i.e. $2 \leftarrow 1 \rightarrow 3$) defined by

@@

\subsection{Persistent Systems}
\label{sec:persistent_systems}
In section \ref{sec:strongly_causal_graphs} we obtained a converse to
part $(a)$ of proposition \ref{prop:ancestor_properties} via the
notion of a strongly causal graph topology.  In this section, we
complete a converse by adding the additional requirement we refer to
as ``persistence''.

@@

This interpretation of the following persistence condition is that
each node stores some ``memory'' of the past.

@@

\begin{remark}
  In the context of Granger-causality, ``most'' systems should be
  persistent.  In particular, $\mathsf{\VAR}(p)$ models are likely to
  be persistent since these naturally result in an equivalent
  $\mathsf{MA}(\infty)$ representation, except for the pathological case
  where $\B(z)$ is Nilpotent.  Indeed, in a $\VAR(1)$ model, if the
  system matrix is diagonalizable and has atleast two distinct
  eigenvalues, then the system is persistent.  See Example
  \ref{ex:persistent_system}.

>>>>

% Should I do this?  It makes cross references much easier,
% but is possibly extra complication.
% \counterwithin{theorem}{section}
% \counterwithin{corollary}{section}
% \counterwithin{proposition}{section}
% \counterwithin{lemma}{section}

% \counterwithin{remark}{section}
% \counterwithin{definition}{section}
% \counterwithin{example}{section}
% \counterwithin{assumptions}{section}

@@

\abstract{We study Granger-causality in the context of wide-sense
  stationary time series, where our focus is on the topological
  aspects of the underlying causality graph.  We establish sufficient
  conditions (in particular, we develop the notion of a ``strongly
  causal'' graph topology) under which the true causality graph can be
  recovered via pairwise causality testing alone, and provide examples
  from the gene regulatory network literature suggesting that our
  concept of a strongly causal graph may be applicable to this field.
  We implement and detail finite-sample heuristics derived from our
  theory, and establish through simulation the efficiency gains (both
  statistical and computational) which can be obtained (in comparison
  to LASSO-type algorithms) when structural assumptions are met.}

@@

\section{Introduction and Review}
\label{sec:introduction}
In this paper we study the notion of Granger-causality
\cite{granger1969investigating} \cite{Granger1980329} as a means of
uncovering an underlying causal structure in multivariate time series.
Though the underlying causality graph cannot be observed directly,
it's it's presence is inferred as a latent structure among observed
time series data.  This notion is leveraged in a variety of
applications e.g. in Neuroscience as a means of recovering
interactions amongst brain regions \cite{bressler2011wiener},
\cite{anna_paper2008}, \cite{david2008identifying}; in the study of
the dependence and connectedness of financial institutions
\cite{NBERw16223}; gene expression networks \cite{Fujita2007},
\cite{methods_for_inferring_gene_regulatory_networks_from_time_series_expression_data},
\cite{grouped_graphical_granger_modelling_for_gene_expression_regulatory_networks_discovery},
\cite{discovering_graphical_Granger_causality_using_the_truncating_lasso_penalty};
and power system design \cite{Misyrlis2016450}, \cite{yuan2014root}.

Granger-causality can generally be formulated by searching for the
``best'' graph structure consistent with observed data, which is in
general an extremely challenging problem (i.e. it may be framed as a
best subset selection problem, see, \cite{hastie_bss_comp} for recent
improvements in BSS methods), moreover, the comparison of quality
between different structures, and hence the notion of ``best'' needs
qualification.  In applications where we are interested merely in
minimizing the mean squared error of a linear one-step-ahead
predictor, then we will be satisfied with an entirely dense graph of
connections, since each edge can only serve to reduce estimation
error.  However, since the number of edges scales quadratically in $n$
(the number of nodes) it becomes imperative to infer a sparse
causality graph for large systems, both to avoid overfitting observed
data, as well as to aid the interpretability of the results.

A fairly early approach to the problem in the context of large systems
is provided by \cite{bach2004learning}, where the authors apply a
local search heuristic to the Whittle likelihood with an AIC
penalization.  The local search heuristic is a common approach to
combinatorial optimization due to it's simplicity, but is liable to
get stuck in shallow local minima.

A second and wildly successful heuristic is the LASSO regularizer
\cite{tibshirani1996regression}, which can be understood as a natural
convex relaxation to penalizing the count of the non-zero edges.  The
LASSO enjoys fairly strong theoretical guarantees
\cite{wainwright2009sharp}, extending largely to the case of
stationary time series data with a sufficiently fast rate of
dependence decay \cite{basu2015} \cite{wong2016lasso}
\cite{autoregressive_process_modelling_via_the_lasso_procedure}, and
variations on the LASSO have been applied in a number of different
time series contexts as well as Granger-causality
\cite{DBLP:journals/corr/HallacPBL17} \cite{haufe2008sparse}
\cite{bolstad2011causal} \cite{he2013stationary}
\cite{grouped_graphical_granger_modelling_for_gene_expression_regulatory_networks_discovery}.
One of the key improvements to the original LASSO algorithm is the
adaptive (i.e. weighted) ``adaLASSO'' \cite{adaptive_lasso_zou2006},
for which oracle results (i.e. asymptotic support recovery) are
established under less restrictive conditions than for the vanilla
LASSO.

% We study a novel perspective on the problem by establishing conditions
% under which simple \textit{pairwise} Granger-causality tests are
% sufficient to recover the entire graph.  Without particular
% topological assumptions, pairwise testing naturally results in a large
% number of false positives \cite{tam2013gene_pwgc}.

In the context of time series data, sparsity assumptions remain
important, but there is significant additional structure that may
arise as a result of considering the topology of the underlying
Granger-causality graph.  The focus of this paper is to shed light on
some of these topological questions, in particular, we study a
particularly simple notion of graph topology which we term
``strongly causal'' and show that stationary times series whose
underlying causality graph has this structure satisfy natural
intuitive notions of ``information flow'' through the graph.
Moreover, we show that such graphs are perfectly recoverable with only
\textit{pairwise} Granger-causality tests, which would otherwise
suffer from serious confounding problems (see
e.g. \cite{tam2013gene_pwgc} for earlier work on pairwise testing).
Aside from being an interesting theoretical perspective, prior
assumptions about the underlying graph (similarly to sparsity
assumptions) can greatly improve upon the statistical power of
causality graph recovery algorithms when the assumptions are met.

@@

The principle contributions of this paper are as follows: firstly, in
section \ref{sec:theory} we study \textit{pairwise} Granger-causality
relations, providing novel theorems connecting the structure of the
causality graph to the pairwise ``causality flow'' in the system, as
well as an interpretation in terms of the graph topology of the
sparsity pattern of matrices arising in the Wold decomposition,
generalizing in some sense the notion of ``feedback-free'' processes
studied by \cite{caines1975feedback} in close connection with
Granger-causality.  We establish sufficient conditions (sections
\ref{sec:strongly_causal_graphs}, \ref{sec:persistent_systems}) under
which a fully conditional Granger-causality graph can be recovered
from pairwise tests alone (sec \ref{sec:pairwise_algorithm}).  We
report a summary of simulation results in \ref{sec:simulation}, with
additional results reported in the supplementary material
\ref{apx:simulation}.  Our simulation results establish that there is
significant potential for improvement over existing methods, and that
the graph-topological aspects of time series analysis are relevant for
both theory and practice.  Concluding remarks on further open problems
and extensions are provided in Section \ref{sec:conclusion}.  The
proofs of each proposition and theorem are also relegated to the
supplementary material, simple corollaries have proofs included in the
main text.

@@

\begin{equation}
  \label{eq:hilbert_space_defn}
  \begin{aligned}
    \H_t^x &= \cl \{\sum_{\tau = 0}^p a_\tau^\T x(t - \tau)\ |\ a_\tau \in \R^n, p \in \N\} \subseteq L_2(\Omega)\\
    H_t^x &= \{a x(t)\ |\ a \in \R\} \subseteq L_2(\Omega),
  \end{aligned}
\end{equation}

where the closure is naturally in mean-square.  We will often omit the
superscript $x$ which should be clear from context.  Evidently these
spaces are separable, and as closed subspaces of a Hilbert space they
are themselves Hilbert.  We will denote the spaces generated in
analogous ways by particular components of $x$ as e.g.
$\H_t^{(i, j)}$, $\H_t^{i}$ or by all but a particular component as
$\H_t^{-j}$.

@@

where $c(t)$ is a purely deterministic sequence, $v(t)$ is an
uncorrelated sequence and $A(0) = I$.  We will assume that $c(t) = 0$.
Given our setup, this representation can be inverted to yield the
$\VAR(\infty)$ form

@@

\begin{definition}[Granger Causality]
  \label{def:granger_causality}
  For the WSS series $x(t)$ satisfying the assumptions of Definition
  \ref{def:basic_setup} we will say that component $x_j$
  \textit{Granger-Causes} (GC) component $x_i$ (with respect to $x$)
  and write $x_j \gc x_i$ if

@@

where $\xi[x \ |\ \H] \defeq \E (x - \linE{x}{\H})^2$ is the mean
squared estimation error and $\linE{x}{\H} \defeq \text{proj}_{\H}(x)$
denotes the (unique) projection onto the Hilbert space $\H$.
\end{definition}

@@

@@

  A \textit{cycle} is a path of length $2$ or more between a node and
  itself.  An edge between a node and itself $(i, i) \in \gcge$ (which
  we do not consider to be a cycle) is referred to as a \textit{loop}.

@@

  Recall that we do not allow a node to be it's own parent, although
  unless $\gcg$ is a DAG, a node \textit{can} be it's own ancestor.
  We will occasionally need to explicitly exclude $i$ from $\anc{i}$,
  in which case we will write $\anc{i}\setminus \{i\}$.
\end{definition}

@@

The following propositions are essentially lemmas used for the proof
of the upcoming Proposition \ref{prop:ancestor_properties}, but
remain relevant for providing intuitive insight into the problems at
hand.

@@

  then $\H_t^{(i)} \perp \H_t^{(j)}$, that is,
  $\forall s, \tau \in \Z_+\ \E[x_i(t - s)x_j(t - \tau)] = 0$.  Moreover,
  this means that $j \npwgc i$ and $\linE{x_j(t)}{\H_t^i} = 0$.
\end{proposition}

@@

\begin{remark}
  It is possible for components of $x(t)$ to be correlated at some
  time lags without resulting in pairwise causality.  For instance,
  the conclusion $j \npwgc i$ of Proposition
  \ref{prop:separated_ancestor_uncorrelated} will still hold even if
  $i \in \anc{j}$, since $j$ cannot provide any information about $i$
  that is not available from observing $i$ itself.
  % However, it is important to note that this is the case only if we
  % have direct access to $x(t)$.  That is, if we instead observed a
  % noisy version of $x$ (i.e. $\tilde{x}(t) = x(t) + n(t)$) the
  % observation of $\tilde{x}_j$ would enable conditionally improved
  % estimates of $\tilde{x}_i$, without there necessarily being any
  % causal flow in the underlying system $x(t)$.
\end{remark}

@@

\begin{definition}[Confounder]
  A node $k$ will be referred to as a \textit{confounder} of nodes
  $i, j$ (neither of which are equal to $k$) if
  $k \in \anc{i} \cap \anc{j}$ and there exists a path
  $\gcgpath{k}{i}$ not containing $j$, and a path $\gcgpath{k}{j}$ not
  containing $i$. A simple example is furnished by the ``fork'' graph
  $i \leftarrow k \rightarrow j$.
\end{definition}

@@

It seems reasonable to expect a converse of Proposition
\ref{prop:ancestor_properties} to hold, i.e.
$j \in \anc{i} \Rightarrow j \pwgc i$.  Unfortunately, this is not the
case in general, as different paths through $\gcg$ can lead to
cancellation (see Figure \ref{fig:diamond_cancellation}).  In fact, we
do not even have $j \in \pa{i} \Rightarrow j \pwgc i$ (see Figure
\ref{fig:lag_cancellation}).

@@

Examples of strongly causal graphs include directed trees (or
forests), DAGs where each node has at most one parent, and Figure
\ref{fig:example_fig3} of this paper.  A complete bipartite graph with
$2n$ nodes is also strongly causal, demonstrating that the number of
edges of such a graph can still scale quadratically with the number of
nodes.  It is evident that the strong causal property is inherited by
subgraphs.

\begin{example}
  Though examples of SCGs are easy to construct in theory, should
  practitioners expect SCGs to arise in application?  While a positive
  answer to this question is not \textit{necessary} for the concept to
  be useful, it is certainly sufficient.  Though the answer is likely
  to depend upon the particular application area, examples appear to
  be available in biology, in particular, the authors of
  \cite{discovering_graphical_Granger_causality_using_the_truncating_lasso_penalty}
  cite an example of the so called ``transcription regulatory network
  of \textit{E.coli}'', and
  \cite{learning_genome_scale_regulatory_networks} study a much larger
  regulatory network of \textit{Saccharomyces cerevisiae}.  These
  networks, which we reproduce in Figure \ref{fig:gene_networks},
  appear to have at most a small number of edges which violate the
  strong-causality condition.

@@

    {\footnotesize Figure \ref{fig:gene_network1} has only one edge
      violating the strong-causality assumption, and Figure
      \ref{fig:gene_network2} appears qualitatively to be nearly
      strongly causal.  In particular, most of the edges are emanating
      from network hubs, and many of the other edges are colliders.

      Figure \ref{fig:gene_network1}
      is reproduced under the Creative Commons Attribution
      Non-Commercial License
      (http://creativecommons.org/licenses/by-nc/2.5) and Figure
      \ref{fig:gene_network2} under the Creative Commons Attribution
      License (https://creativecommons.org/licenses/by/4.0/)}
  \end{figure}
\end{example}

For later use, and to get a feel for the topological implications of
strong causality, we explore a number of properties of such graphs
before moving into the main result of this section.  The following
important property essentially strengthens Proposition
\ref{prop:ancestor_properties} for the case of strongly causal graphs.

@@

\begin{corollary}
  \label{cor:parent_corollary}
  If $\gcg$ is a strongly causal DAG then $i \pwgc j$ and $j \in \anc{i}$ are
  \textit{alternatives}, that is $i \pwgc j \Rightarrow j \notin \anc{i}$.
\end{corollary}
\begin{proof}
  Suppose that $i \pwgc j$ and $j \in \anc{i}$.  Then since $\gcg$ is
  acyclic $i \not\in \anc{j}$, and by Proposition
  \ref{prop:ancestor_properties} there is some
  $k \in \anc{i}\cap\anc{j}$ which is a confounder.  However, by
  Proposition \ref{prop:sc_graph_common_anc} $k$ cannot be a
  confounder, a contradiction.
\end{proof}
% This is a direct proof
  % \begin{proof}
%   Suppose that $\gcg$ is a strongly causal DAG and that we have both
%   $i \pwgc j$ and $j \in \anc{i}$, which implies that
%   $i \not \in \anc{j}$ since $\gcg$ is a DAG.  We will establish the
%   contradiction $i \in \anc{j}$.

%   Since $j \in \anc{i}$ there is a path $\gcgpath{j}{i}$.
%   Moreover, since $i \pwgc j$ by Proposition \ref{prop:pwgc_anc} there
%   must be a confounding ancestor $k \in \anc{i} \cap \anc{j}$, where the
%   node $k$ has a path to $j,\ \gcgpath{k}{j}$, as well as a path
%   to $i,\ \gcgpath{k}{i}$.

%   \hl{Double check that the ancestor properties Proposition implies
%     that BOTH paths do not contain the other node.}

@@

\begin{corollary}
  \label{cor:bidirectional_edge}
  If $\gcg$ is a strongly causal DAG such that $i \pwgc j$ and
  $j \pwgc i$, then $i \not\in \anc{j}$ and $j \not\in \anc{i}$.  In
  particular, a pairwise bidirectional edge indicates the absence of
  any edge in $\gcg$.
\end{corollary}
\begin{proof}
  This follows directly from applying Corollary
  \ref{cor:parent_corollary} to $i \pwgc j$ and $j \pwgc i$.
% This is a direct proof
% \begin{proof}
  % By way of contradiction, suppose that $i \in \pa{j}$.  We will
  % consider the two possibilities allowed by proposition
  % \ref{prop:ancestor_properties} for $j \pwgc i$.  Firstly
  % $j \in \anc{i}$ is impossible since $\gcg$ is assumed to be acyclic.
  % Secondly, if there is some confounding node
  % $u \in \anc{i} \cap \anc{j}$ with a path
  % $u \rightarrow \cdots \rightarrow j$ which does not contain $i$ we
  % have a contradiction since there must now be multiple
  % $u \rightarrow j$ paths: the aforementioned, and a path
  % $u \rightarrow \cdots \rightarrow i \rightarrow \cdots \rightarrow
  % j$ which \textit{does} contain $i$.  We conclude that $i \in \pa{j}$
  % is impossible, and symmetrically that $j \in \pa{i}$ is as well.
% \end{proof}
\end{proof}

In light of Proposition \ref{prop:sc_graph_common_anc}, the following
provides a partial converse to Proposition
\ref{prop:ancestor_properties}, and supports the intuition of ``causal
flow'' through paths in $\gcg$.

@@

\begin{remark}
  As we have seen and as is true in much of statistics, confounding
  nodes pose challenges for Granger-causality.  However, as opposed to
  the causal calculus of \cite{pearl2009causality}, pairwise
  Granger-causality does not suffer any difficulty with so-called
  ``colliders'', that is, the topology $i \rightarrow k \leftarrow j$ will never result
  in $i \pwgc j$ or $j \pwgc i$.  This is evidently an advantage of
  the \textit{temporal} nature of Granger-causality -- there is no
  backwards causal flow along the edges of $\gcg$.
\end{remark}

\begin{example}
  As a final remark of this subsection we note that a complete
  converse to Proposition \ref{prop:ancestor_properties} is not
  possible without additional conditions.  Consider the ``fork''
  system on $3$ nodes (i.e. $2 \leftarrow 1 \rightarrow 3$) defined by

@@

\subsection{Persistent Systems}
\label{sec:persistent_systems}
In section \ref{sec:strongly_causal_graphs} we obtained a converse to
part $(a)$ of Proposition \ref{prop:ancestor_properties} via the
notion of a strongly causal graph topology.  In this section, we
complete a converse by adding the additional requirement we refer to
as ``persistence''.

@@

The interpretation of the following persistence condition is that
each node stores some ``memory'' of the past.

@@

\begin{remark}
  In the context of Granger-causality, ``most'' systems should be
  persistent.  In particular, $\mathsf{\VAR}(p)$ models are likely to
  be persistent since these naturally result in an equivalent
  $\mathsf{MA}(\infty)$ representation, except for the pathological case
  where $\B(z)$ is Nilpotent.  Indeed, in a $\VAR(1)$ model, if the
  system matrix is diagonalizable and has at least two distinct
  eigenvalues, then the system is persistent.  See Example
  \ref{ex:persistent_system}.
